Example workflow / tutorial for the seegSDM package
========================================================

This document runs through a typical workflow for distribution modelling using the ```seegSDM``` package.

It will include importing and checking data, running BRT ensembles in parallel and examining the fitted models.
This is a work in progress, so sorry if it stops halfway through or doesn't make any sense yet! Please report any issues via the [issues reporting system](https://github.com/SEEG-Oxford/seegSDM/issues).

The structure is:

##### [Loading data](#load)
##### [Quality control](#quality)
##### [Generating pseudo-absences](#pseudo)
##### [Running a single BRT model](#BRT)
##### [Running a BRT ensemble in parallel](#ensemble)
##### [Visualising the BRT ensemble](#vis)
##### [Outputting the results](#output)



### <a id="load"></a>Loading data

First we load in some occurrence data. Here we'll use fake occurrence data from a csv file on the seegSDM github page.

```{r}
summary(cars)
```

### <a id="quality"></a>Quality control

```{r}
summary(cars)
```

### <a id="pseudo"></a>Generating pseudo-absences

```{r}
summary(cars)
```

### <a id="BRT"></a>Running a single BRT model 

```{r}
summary(cars)
```

### <a id="ensemble"></a>Running a BRT ensemble in parallel

```{r}
summary(cars)
```

### <a id="vis"></a>Visualising the BRT ensemble

```{r}
summary(cars)
```
### <a id="output"></a>Outputting the results

```{r}
summary(cars)
```

